The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2018-10-26 Paper/jaccardbaetica.tex
add some lines consumption cent

The majority of olive oil amphorae produced in this province and shipped through the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti} with different information. However, inscriptions and \textit{tituli picti} have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae. During almost three centuries, stamps were used to mark amphorae with a different chronological frequency \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

%Many hypothesis involved the meaning of stamps. 
 There is not a general consensus about the meaning of the stamps \citep{rodriguez_baetican_1998}. It seems that stamps could have been identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could belong to the owner of the making amphorae workshops or even a production counting system  \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this province was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of Baetica province in the Roman organization. Indeed, archaeological evidence shows a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

%hipotesis: si esto es así entonces es asá
Here, therefore, this study proposes an evolutionary approach to explore the effect of Baetican olive oil production by computing of dissimilarity index. The assumption of this analysis is that closer workshops concentrate similar amphoric stamps in a specific area than the distant workshops. By contrast, if we found similar amphoric stamps in different areas between workshops then the correlation between spatial distance and workshops would not be valued.

@@

\subsection{Case study}

Our case study examines the relation between the distribution of amphoric stamps and the workshops. The amphora workshops were situated in different locations in Baetica province, along the river Guadalquivir and its tributary Genil in order to detect similarities between stamps from workshops and spatial distance (see Fig.\ref{workshop})

@@

\subsection{Quantifying the cultural diversity}

>>>>

The majority of olive oil amphorae produced in this province and shipped through the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti} with different information. However, inscriptions and \textit{tituli picti} have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae.
The fascination aroused by stamps has caused it to be one of the inscriptions with the most publications about its origin and meaning. Its endure in time become it an essential element that were used to mark amphorae during almost three centuries \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

%Many hypothesis involved the meaning of stamps. 
There is not a general consensus about the meaning of the stamps \citep{rodriguez_baetican_1998}. The meaning of these stamps could be identified based on two premises: content (olive oil) and context (amphora workshop). On the one hand, it seems that stamps could have been identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other, they could belong to the owner of the making amphorae workshops or even a group of amphora workers (CITAR). Additionally, they have been related to a quality system \citep{callender} or a productin counting system \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this production was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of Baetica province in the Roman organization. Indeed, archaeological evidence shows a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

%hipotesis: si esto es así entonces es asá
Here, therefore, this study proposes an evolutionary approach to explore the effect of Baetican olive oil production by computing the distribution of stamps. To do that, three case study have been studied to analyse the relation between production centres (Baetica) and consumption centres (Britannia and Germania).

The assumption of this study is analysing the effect of the production patterns between different centres.  This effect could affect due to different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamp in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province. In the case of Baetica province, we want to identify the role of the stamps in the organization of the workshop. In Roman provinces, our aim is detecting groups of stamps concentrated in a area or if some groups have an important role for the exportation of olive oil in those provinces.

@@

\subsection{Production centres: Baetica province}

Our case study examines the relation between the distribution of amphoric stamps and the workshops. In particular, we study the distribution of amphoric stamp with the aim of identifying a correlation between geographical distance and amphoric stamps. Based on this assumption, we proposed three hypotheses: a) it exists a correlation between spatial distance and the distribution of stamps, b) stamps located in close workshops share similar traits and c) Low mobility of amphoric stamps to other region: stamps always stay in the same region.

The amphora workshops were situated in different locations in Baetica province, along the river Guadalquivir and its tributary Genil in order to detect similarities between stamps from workshops and spatial distance (see Fig.\ref{workshop}).

@@

\subsection{Consumption centres}

The conquest of new provinces allowed to the Roman Empire the arrival of new resources. This lead to a progressive change in the economic and social structure. As a result, Augusto's administration created the figure of the \textit{praefectura annonae} for the supply of wheat. The role of the praefectura annonae was mainly to provide though \textit{frumentationes} a quantity per month of wheat to Roman citizens CITAR (Remesal, 1986; 2011).

\subsection{Britannia}
